Value-Suppressing Uncertainty Maps

### 2017-03-31 paper/template.tex
Split up conditions figure, scoped back up from thematic maps to vis generally

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in thematic maps. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Compared to a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. \figref{fig:example} shows examples of both a more traditional bivariate map and a VSUM.

@@

There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots~\cite{hintze1998violin} use a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps. To our knowledge, there are no established standards of visualizing uncertainty in thematic maps (but see MachEachren~\cite{maceachren1992visualizing} for a glossary of potential techniques).

@@

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In thematic maps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have more freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background) and size. Correll \& Gleicher~\cite{correll2013error} show that even audiences without statistical backgrounds can interpret uncertainty information encoded in these channels.

@@

We performed a crowdsourced experiment on Amazon's Mechanical Turk to evaluate the effectiveness of VSUMs for integrating uncertainty and value information in thematic maps. This focus on integration meant that we limited our experimental tasks to scenarios where the participants needed to consider both value and uncertainty before making a decision. We gave participants two main tasks:

@@

For each task, participants saw one of three types of thematic maps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{2D bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). All factors were presented within subjects. Secondary factors were evenly blocked across stimuli. Participants reported fatigue effects in piloting, and so we limited the number of replicates in this study: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.

>>>>

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in visualizations. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Compared to a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. \figref{fig:example} shows examples of both a more traditional bivariate map and a VSUM.

@@

There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots~\cite{hintze1998violin} use a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps. To our knowledge, there are no established standards of visualizing uncertainty in heatmaps.

@@

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In heatmaps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have more freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background) and size. Correll \& Gleicher~\cite{correll2013error} show that even audiences without statistical backgrounds can interpret uncertainty information encoded in these channels.

@@

We performed a crowdsourced experiment on Amazon's Mechanical Turk to evaluate the effectiveness of VSUMs for integrating uncertainty and value information in visualizations. This focus on integration meant that we limited our experimental tasks to scenarios where the participants needed to consider both value and uncertainty before making a decision. We gave participants two main tasks:

@@

For each task, participants saw one of three types of heatmaps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{2D bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). All factors were presented within subjects. Secondary factors were evenly blocked across stimuli. Participants reported fatigue effects in piloting, and so we limited the number of replicates in this study: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.
